---
title: 'Practical Machine Learning: Activity Prediction'
author: "Travis May"
date: "Saturday, November 22, 2014"
output: html_document
---

## Summary
The intent of this project will be to use data from accelerometers on the belt,
forearm, arm, and dumbell of 6 participants.

Participants were asked to perform barbell lifts correctly and incorrectly in 5
different ways. More information is available from the website here:
http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Data
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible
to collect a large amount of data about personal activity relatively inexpensively.
These type of devices are part of the quantified self movement of a group of
enthusiasts who take measurements about themselves regularly to improve their health,
to find patterns in their behavior, or because they are tech geeks.
One thing that people regularly do is quantify how much of a particular activity they do,
but they rarely quantify how well they do it.

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.

## Analysis
Before the analysis can be done in earnest, a few steps are required.

### Initial Work
Initially, the environment's working directory must be pointed to the correct
local directory.
```{r, echo=FALSE}
setwd("~/datasciencecoursera/prac_mech")
```

Next, all required libraries must be loaded.
```{r}
library(caret)
library(foreach)
library(doSNOW)
library(data.table)
library(knitr)
```
To facilitate the computing process, parallel processing is required.
```{r, echo=FALSE}
cl <- makeCluster(4, type="SOCK")
registerDoSNOW(cl)
```

### Download Data
```{r}
download <- function() {

  trainURL <-"http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
  testURL <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

  download.file(trainURL, "training.csv")
  download.file(testURL, "testing.csv")
}

download()
```
### Read the files into the environment
```{r}
read <- function(file) {
  fread(file, na.strings=c("#DIV/0!",""))
}

trainRaw <- read("training.csv")
testRaw <- read("testing.csv")
```

### Set the seed for prediction
```{r}
set.seed(1986)
```

### Cleaning the data
Any columns that contain incomplete data need to be removed from the data frames.
Additionally, the training data frame has a column named 'classe' which needs
to be converted to the factor type.

```{r}
## Drops unnecessary columns
drop <- function(x) {
  return(x[,!c("V1", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2",
  "cvtd_timestamp", "new_window", "num_window"),with=F])
}

## Convert 'classe' column into factor type for processing.
transformFeatures <- function(x) {
  return(x[,classe:=factor(classe)])
}

filter <- function(df) {

  # Convert name of data frame into string to test for features in data.
  frameName <- deparse(substitute(df))

  # Remove any rows that contain NA values
  missing <- sapply(df, function(x) any(is.na(x)))
  df <- df[,eval(names(which(missing == FALSE))),with=F]

  # Drop unnecessary columns
  df <- drop(df)

  # Transform training-data-specific "classe" column into factor type.
  if (grepl("^(train)",frameName) == TRUE) {
    if ("classe" %in% colnames(df)) {
      transformFeatures(df)
    }
  } else if (grepl("^(test)", frameName) == TRUE) {
    #continue
  } else {
    print("Check that the data frames are correctly labeled.")
  }

  return(df)
}
```

Using the previous functions, the data can be effectively cleaned.
```{r}
trainFeatures <- filter(trainRaw)
testFeatures <- filter(testRaw)
```

### Building Models and Testing Predictions
Three types of models will be used: 'Random Forest', 'SVM', also known as
'Radial Kernel', and 'KNN'.

The control uses 5-fold cross validation.
```{r}
control <- trainControl(method = "cv", number = 5, allowParallel = TRUE,
                        verboseIter = TRUE)
model1 <- train(classe ~ ., data=trainFeatures, method = "rf",
                trControl = control)
model2 <- train(.outcome ~ ., data=trainFeatures, method = "svmRadial",
                trControl = control)
model3 <- train(.outcome ~ ., data=trainFeatures, method="knn",
                trControl = control)
```

```{r, echo = FALSE}
accuracy <- data.frame(Model=c("Random Forest", "SVM (radial)", "KNN"),
                      Accuracy=c(round(max(head(model1$results)$Accuracy), 3),
                                 round(max(head(model2$results)$Accuracy), 3),
                                 round(max(head(model3$results)$Accuracy), 3)))
```
Below, one can see the accuracy of each model side-by-side.
```{r, results='asis'}
kable(accuracy)
```

Now, predictions can be derived from each model.
```{r, echo = FALSE}
prediction1 <- predict(model1, testFeatures)
prediction2 <- predict(model2, testFeatures)
prediction3 <- predict(model3, testFeatures)
```

For comparison, a table with the prediction values will be created.
```{r}
predictions <- data.frame(rf = prediction1, svm = prediction2,
                          knn = prediction3)
predictions$agree <- with(predictions, rf == svm && rf == knn)
agree <- all(predictions$agree)

colnames(predictions) <- c("Random Forest", "SVM", "KNN", "Agreement")
```

All models agree.
```{r,results='asis'}
kable(predictions)
```

### Submission
The files can now be created for grading.
```{r}
answers <- predictions[,1]

pml_write_files <- function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

## Write files for submission
pml_write_files(answers)
```
